---
title: "Geog4/6300: Lab 1"
format: gfm
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)

knitr::opts_chunk$set(echo = TRUE)
```

## Loading data into R, data transformation, and summary statistics

**Your name:**
Wil Gamble
**Overview and lab criteria:**

This lab is intended to assess your ability to use R to load data and to generate basic descriptive statistics. For this lab to be marked complete, the following criteria must be met:

4. Identify and apply appropriate data filtering and cleaning strategies to prepare datasets for analysis. (Task 2)
5. Effectively interpret the code you create, explaining in plain language what each step does to the data. (Task 6)
6. Identify and use appropriate external documentation — including package references, help files, and peer resources — to learn and apply unfamiliar functions or methods. (Task 7)
7. Filter, aggregate, and transform datasets using grouping and summary operations to answer specific analytical questions. (Tasks 1 & 5)
9. Reshape data between wide and long formats to meet the requirements of different analytical or visualization tasks. (Task 4)
11. Create effective data visualizations across multiple chart types (line, scatter, histogram, Q-Q plot), applying appropriate aesthetic choices such as color, grouping, and labeling. (Task 3 & 5)

**Data:**

You'll be using monthly weather data from the Daymet climate database (http://daymet.ornl.gov) for all counties in the United States over an 12-year period (2010-2021). These data are available on the GitHub repo for our course. The following variables are provided:

* `cty_txt`: Code for joining to census data
* `year`: Year of observation (with an initial "Y" to make it a character)
* `month`: Month of observation (1 = Jan, 2 = Feb, etc.)
* `median_tmax`: Median maximum recorded temperature (Celsius)
* `median_tmin`: Median minimum recorded temperature (Celsius)
* `sum_prcp`: Total recorded precipitation for the month (mm)
* `cty_name`: Name of the county
* `state`: State of the county
* `region`: Census region (map: https://www2.census.gov/geo/pdfs/maps-data/maps/reference/us_regdiv.pdf)
* `division`: Census division
* `X`: Longitude of the county centroid
* `Y`: Latitude of the county centroid

These labs are meant to be done collaboratively, but your final submission should demonstrate your own original thought (don't just copy your classmate's work or turn in identical assignments). Your answers to the lab questions should be typed in this Quarto template. You'll then render the document to a GitHub markdown document and upload it to your class GitHub repo.

**Procedure:**

Load the tidyverse package and import the data:

```{r}
#| message: false

library(tidyverse)

daymet_data <- read_csv("data/daymet_monthly_median_2010-2021.csv")
```

We can look at the first few rows of the dataset using the *head()* function. We also use *kable* to format the output as a readable table..

```{r}
kable(head(daymet_data))
```

There are a lot of observations here, 452,448 to be exact. To get a better grasp on the data, we can use `group_by()` and `summarise()` from the tidyverse package. This will allow us to identify the mean value for each year by county across the study period.

## Task 1

*Use `group_by()` and `summarise()` to calculate the mean minimum temperature for each year by county across all months, also including State and Region as grouping variables. Your resulting dataset should show the value of tmin for each county in each year. Use the `kable()` and `head()` functions as shown above to call the resulting table.*

```{r}
#| label: task1

Min_Temp_Daymet<-daymet_data%>%
  group_by(state,region,cty_name,year)%>%
  summarise(median_of_tmin=mean(median_tmin))


kable(head(Min_Temp_Daymet))

# Your code goes here
```

## Task 2

*Let's shift to the state level, focusing on those in the South Region. Filter the original data frame (`daymet_data`) to just include counties in this region. Then calculate the mean minimum temperature by year for each state. For an optional extra challenge, use the `round()` function to include only 1 decimal point. Use `kable()` and `head()` to call the first few lines of the resulting table.*

```{r}
#| label: task2

South_Region_tmin_mean<-daymet_data%>%
  filter(region=="South Region")%>%
  group_by(state,region,year)%>%
  summarise(median_of_tmin=mean(median_tmin))

kable(head(South_Region_tmin_mean))

# Your code goes here
```

## Task 3

*To visualize the trends, we could use ggplot to visualize change in mean temperature over time. Create a line plot (`geom_line`) showing the state means you calculated in task 2. Use the `color` parameter to show separate colors for each state. You may also need to define the state as a group in the aesthetic parameter.*

```{r}
#| label: task3

ggplot(South_Region_tmin_mean,
       aes(x=median_of_tmin,
           y=state,
           color=state))+
  geom_line()

# Your code goes here
```

## Task 4

*If you wanted to look at these data as a table, you'd need to have it in wide format. Use the `pivot_wider()` function to create a wide-format version of the data frame you created in task 2. In this case, the rows should be states, the columns should be the years, and the data in those columns should be mean minimum temperatures. Then call the whole table using `kable()`.*

```{r}
#| label: task4

South_Region_tmin_mean_Wide<-South_Region_tmin_mean%>%
  pivot_wider(names_from=state,
              values_from=median_of_tmin)
  

kable(South_Region_tmin_mean_Wide)
  

# Your code goes here
```

## Task 5

*Let's assess the relationship of heat and precipitation by region. Returning to the original dataset, create a data frame that shows the mean maximum temperature and mean precipitation for all states in 2015, also including region as a subgroup in your `group_by`. Then use ggplot to create a scatterplot (`geom_point`) for these two variables, coloring the points using the region variable.*

```{r}
#| label: task5

Mean_Temp_Precip<-daymet_data%>%
  filter(year=="Y2015")%>%
   group_by(state,region)%>%
  summarise(mean_of_tmax=mean(median_tmax), 
            mean_prcp=mean(sum_prcp))

ggplot(Mean_Temp_Precip,
       aes(x=mean_of_tmax,
           y=mean_prcp,
         color=region))+
  geom_point()


# Your code goes here
```

## Task 6

*In the space below, explain what each function in your code for task 5 does to the dataset in plain English.*

The first function of the code is creating a new object that will transfer alterations to a new name.
That is Mean_Temp_Precip<-daymet_data. The %<% is a "pipe". I understand it as a separation for a new function 
to be run. It gives the code a sequence of events. filter() is used to denote what from a column we want to keep and be used in our data set. You start with the column header then equal it to the column variable you want. group_by() takes a whole column and transfers it to the new object. summarize() allows you to create a new column/data frame that at the same time can perform a function like average, mean, etc. summarise(median_of_tmax=mean(median_tmax) creates a new column called "median_of_tmax" based on mean data from the existing column "median_tmax".   

ggplot() is used to create/alter different graphs. You start with the newly (or any) data set created. The aes function can then be used to tell ggplot what columns are going to control the graph. You can then decide the x/y axis using x=, y=. The color= function will make the data of a certain column different colors. Ones that share a name will be the same color. geom_point() is what visualizes all that is above into a scatter plot. There are multiple geom_blanks() that provide different graphs. 


*Your response here.*

## Task 7

*The `dplyr` package also includes `across` function. Use `?across` on the R command line to open the documentation for this function. In the space below, explain what it does in your own words. Then interpret the way the across function is used below, going line by line within the function.*

```{r}
state_2015 <- daymet_data %>%
  filter(year == "Y2015") %>%
  group_by(region, state) %>%
  summarise(
    across(
      c(median_tmax, sum_prcp),
      mean,
      na.rm = TRUE,
      .names = "mean_{.col}"
    )
  )
```

{Your response here}

The across function appears to remove the need to type out the new individual names and mean function each time. The across(
      c(median_tmax, sum_prcp),
      mean 
is making the code take the mean of the existing columns named in c().
      na.rm = TRUE
prevents na values from preventing the formation of columns properly. It will tell R to ignore NA values. This allows the mean calculation to be performed without interference. 
      .names = "mean_{.col}
adds mean_ to the start of median_tmax, and sum_prcp. For just 2 columns, it is not too useful here. If I needed to deal with 50 columns, I could see the value.

## Challenge Question

In class, we covered ways of working with the Daymet API. Create a script below that uses the **daymetr** package to download data from Daymet for a place (or places) of your choosing. Then visualize the temporal pattern for a variable of your choosing in this place, similar to what you did in question 4. Use a dplyr function (`mutate()`, `summarise()`, `filter()`, etc.) to do any needed data wrangling and create a visual using ggplot.

In addition to this code, write a short summary of a pattern that's evident in the data you visualized.

```{r}
#| label: challenge
library(daymetr)

terrell_daymet<-download_daymet(site="terrell County, Ga",lat=31.77,lon=-84.43,start=2010,end=2020,internal=TRUE)

Terrell_data<-terrell_daymet$data


Terrell_Mean_Mtemp_precip_2016<-Terrell_data%>%
  filter(year=="2016")%>%
   select(yday,tmax..deg.c.,prcp..mm.day.)

ggplot(Terrell_Mean_Mtemp_precip_2016,
       aes(x=tmax..deg.c.,
           y=prcp..mm.day.,
         color=tmax..deg.c.,
         ))+
  geom_line()
  
# Your code goes here
```

*Explanation goes here.*

For my graph, I wanted to see if there was a pattern between hotter temperatures and more precipitation in 2016. Really it ended up showing that 1. Terrell is either boom or bust for precipitation and 2. It rains more in the fall/spring than the summer/winter. Most of the rainfall is concentrated in the middle of the graph which is where temperatures are at their least extreme. The highest rains are around 66-80 degrees which makes sense. That is the temp range for May. It gets pretty wet then from my memory and a quick Google search corroborated that. 

http://weather.uga.edu/mindex.php?variable=AV&site=DAWSON (Dawson=terrell county)

## Final Submission Stuff

### Disclosure of Assistance

Besides class materials, what other sources of assistance did you use while completing this lab? These can include input from classmates, relevant material identified through web searches (e.g., Stack Overflow), or assistance from ChatGPT or other AI tools. How did these sources support your own learning in completing this lab?

This lab I mainly used ChatGPT, when I popped an error code. My most common mistake was finding a similar chunk in a script but forgetting to place a %>% in my code. It would say for example "object 'median_of_tmin' not found". I scrambled trying to figure it out, but it was just the lack of a pipe. You helped me quite a bit, when I had all the na data in task 4. I just did not follow task 2 correctly which led to a wrong answer down the line. I tried group_by() for the challenge question but it gave me too many variables. CHATgpt told me to use select instead. I now understand group_by() is for keeping summarize together with certain columns. Select just lets you pick out certain columns. 
*Response here.*

### Lab Reflection

How do you feel about the work you did on this lab? Was it easy, moderate, or hard? What are the biggest things you learned by completing it?

This lab was easier than the first to me. I mainly focused on using script 10 and 8 to understand how the different code chunks needed to be formatted. I have better notes now too which helps. I did not have to use much generative A.I this time. If I have a template, I can normally figure out what to do. Some errors would have been a headache to solve without it though.  I would give this one a moderate compared to lab one's hard. Focusing on the scripts helped. 

The challenge question was not to bad once I found what I needed in script 3. It was mostly just taking earlier code after that. 

*Your response here.*